---
title: "Color Writeup"
author: "Terrance Luangrath"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
---

```{r setup, message=FALSE, echo=FALSE, output=FALSE}
library(tidyverse)
library(MASS)

diamond <- read.csv("diamonds4.csv")
```

# Data Description

The data set, `diamonds4.csv`, contains 5 varaibles that describes more than 1000 different diamonds for sale. 

- `Carat`: *fill in description*
- `Clarity`: *fill in description*
- `Color`: how colorless a diamond is, the more colorless the diamonds, the rarer it tends to be
- `Cut`: *fill in description*
- `Price`: the price value of the diamond in USD


## Color

The diamond color refers to how colorless a diamond is. From the [Blue Nile](https://www.bluenile.com/education/diamonds) website, color is the second most important of the 4Cs of diamond. The more colorless the diamond is the rarer it is. Diamond colors are classify in three main categories, 

- Colorless Diamonds
- Near-Colorless Diamonds
- Faint Diamonds

from these three catgories, diamond color are grades from the ranges of D (colorless) to K (faintly colored). However, in this dataset, there exists only diamond graded from D to J.

```{r color-count-plot, message=FALSE, echo=FALSE, fig.width=4, fig.height=3, fig.cap="Diamond Color Distribution"}
ggplot(diamond, aes(x=color, fill = color)) +
  geom_bar() +
  labs(title = "Distribution of Diamond Colors",
       x = "Color",
       y = " Count") +
  theme_minimal() +
  theme(legend.position = )
```

Looking at Figure $\ref{fig:color-count-plot}$, the distribution of diamond color appears fairly consistent, with counts ranging between 150-200 for most color grades. However, color `J` semms to have around 100 diamonds in the data set. 

Each diamond color falls into one of the three categories from Colorless to Faintly Colored. 

- Colorless diamonds: `D`, `E`, `F` Color Diamonds
- Near-colorless diamonds: `G`, `H`, `I`, `J` Color Diamonds
- Faint color diamonds: `K` Color Diamonds

Since there is no `K` in the data set, it would be assumed there is no faint color diamonds in the data set. To facilitate the analysis, we will create a new column called `color_cateogry` to group each `color` with their respected category. 


```{r color-category-pie-chart, message=FALSE, echo=FALSE, fig.width=4, fig.height=3, fig.cap="Diamond Color Category visualize through a Pie Chart"}

diamond <- diamond %>%
  mutate(color_category = case_when(
    color %in% c("D", "E", "F") ~ "Colorless",
    color %in% c("G", "H", "I", "J") ~ "Near-Colorless",
    color %in% c("K") ~ "Faint"
  ))

color_counts <- diamond %>%
  count(color_category)

ggplot(color_counts, aes(x="", y=n, fill=color_category)) + 
  geom_bar(stat="identity", width=1) + 
  coord_polar("y", start=0) + 
  labs(title = "Distribution of Diamond Color Groups",
       x = NULL,
       y = NULL,
       fill = "Color Category") + 
  theme_void()

```

Looking at \ref{fig:color-category-pie-chart}, the distribution of diamond colors seems to be nearly even, with `Colorless` containing slightly higher amount of diamonds. With a well spread of diamond colors, further analysis is needed to understand the claim and how diamond color would be correlates with the other factors in the data set.



## Relationship with other variables


