---
title: "Investigating Blue Nile's Claim on Buying Diamonds"
author: "Humaid Billoo, Katie A. Dunning, Terrance K. Luangrath, Kristina V. Quintana"
output: 
  bookdown::html_document2
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

# Introduction

# Claim

# Data Description

```{r libraries-dataset, message=FALSE, echo=FALSE, output=FALSE, warning=FALSE}
rm(list = ls())
library(tidyverse)
library(MASS)
library(patchwork)
library(kableExtra)
library(knitr)
library(scales)


diamond <- read.csv("diamonds4.csv")
```

```{r sort-label, echo=FALSE, output=FALSE, message=FALSE, error=FALSE, warning=FALSE}
diamond<-diamond%>%
  mutate(cut=cut%>%
           fct_relevel(c("Good","Very Good","Ideal","Astor Ideal")))%>%
  mutate(clarity=clarity%>%
           fct_relevel(c("SI2","SI1","VS2","VS1","VVS2","VVS1","IF","FL")))%>%
  mutate(color=color%>%
           fct_relevel(c("J","I","H","G","F","E","D")))

diamond <- diamond %>%
  mutate(carat_bin = cut(carat,
                         breaks = c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 8),
                         labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1", "1-1.5", "1.5-2", "2-3", "3-4", "4-5", "5-8")),
                         right = FALSE) %>%
  drop_na(carat) 

carat_new_bins <- diamond %>%
  drop_na(carat) %>%
  mutate(carat_bin = cut(carat,
                         breaks = c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 8),
                         labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1", "1-1.5", "1.5-2", "2-3", "3-4", "4-5", "5-8"),
                         right = FALSE))

CutCarat<-carat_new_bins%>%
  group_by(cut,carat_bin)%>%
  summarise(Ave_Price=round(mean(price),0))%>%
  ungroup()

ColorCarat<-carat_new_bins%>%
  group_by(color,carat_bin)%>%
  summarise(Ave_Price=round(mean(price),0))%>%
  ungroup()

ClarityCarat<-carat_new_bins%>%
  group_by(clarity,carat_bin)%>%
  summarise(Ave_Price=round(mean(price),0))%>%
  ungroup()

CutColor<- carat_new_bins %>%
  group_by(cut,color)%>%
  summarise(Ave_Price=round(mean(price),0))%>%
  ungroup()
```


# Univariate Analysis

## Cut

```{r cut-bar-plot, fig.width=3,fig.height=2,fig.align="center",echo=FALSE, fig.cap = "Distribution of diamond counts by cut quality."}
ggplot(diamond, aes(x=cut))+
  geom_bar(fill = "deepskyblue3", color ="black")+
  scale_y_continuous(breaks=seq(0,800,100))+
  labs(x="Cut", y="Counts",
       title="Diamond Cut Distribution") + 
  theme_minimal() + 
  theme(legend.position = "none")
```



## Color
```{r color-count-plot, message=FALSE, echo=FALSE, fig.width=3, fig.height=2, fig.align="center",fig.cap="Distribution of diamond counts by color grade"}
ggplot(diamond, aes(x=color)) +
  geom_bar(fill = "deepskyblue3", color ="black") +
  labs(title = "Diamond Color Distribution",
       x = "Color",
       y = " Counts") +
  theme_minimal() 
```

A random paragraph can also be an excellent way for a writer to tackle writers' block. Writing block can often happen due to being stuck with a current project that the writer is trying to complete. By inserting a completely random paragraph from which to begin, it can take down some of the issues that may have been causing the writers' block in the first place.


```{r color-category-pie-chart, message=FALSE, echo=FALSE, fig.align = "center", fig.width=3, fig.height=2, fig.cap="Distribution of diamond counts by color groups"}

diamond <- diamond %>%
  mutate(color_category = case_when(
    color %in% c("D", "E", "F") ~ "Colorless",
    color %in% c("G", "H", "I", "J") ~ "Near-Colorless",
    color %in% c("K") ~ "Faint"
  ))

color_counts <- diamond %>%
  count(color_category)

ggplot(color_counts, aes(x="", y=n, fill=color_category)) + 
  geom_bar(stat="identity", width=1, color="black") + 
  coord_polar("y", start=0) + 
  labs(title = "Distribution of Diamond Color Groups",
       x = NULL,
       y = NULL,
       fill = "Color Category") + 
  theme_void()
```

A random paragraph can also be an excellent way for a writer to tackle writers' block. Writing block can often happen due to being stuck with a current project that the writer is trying to complete. By inserting a completely random paragraph from which to begin, it can take down some of the issues that may have been causing the writers' block in the first place.


## Clarity

```{r clarity-bar-count, echo=FALSE, fig.width=4, fig.height=2, fig.align = "center", fig.cap = "Distribution of diamond counts by clarity grade"}
# arrange clarity by rating
clarity_ordered <- diamond %>%
  mutate(clarity = factor(clarity, levels = c("SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF", "FL"))) %>%
  arrange(clarity)

# bar chart of counts and clarity grade
ggplot(clarity_ordered, aes(x=clarity))+
  geom_bar(fill = "deepskyblue3", color = "black")+
  labs(x="Clarity Grade", y="Counts", title="Diamond Clarity Distribution") +
  theme_minimal()
```

A random paragraph can also be an excellent way for a writer to tackle writers' block. Writing block can often happen due to being stuck with a current project that the writer is trying to complete. By inserting a completely random paragraph from which to begin, it can take down some of the issues that may have been causing the writers' block in the first place.


## Carat

```{r carat-count-plot, warning=FALSE,message=FALSE,echo=FALSE,fig.width=4,fig.height=2.5, fig.align = "center", fig.cap="Distribution of diamond counts by carat bins"}
ggplot(diamond, aes(x = carat)) +
  geom_histogram(fill = "deepskyblue3", color = "black") + 
  labs(title = "Diamond Carat Distribution",x = "Carat",y = "Counts") +
  scale_x_continuous(breaks = seq(0, max(diamond$carat), by = 0.5)) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```

A random paragraph can also be an excellent way for a writer to tackle writers' block. Writing block can often happen due to being stuck with a current project that the writer is trying to complete. By inserting a completely random paragraph from which to begin, it can take down some of the issues that may have been causing the writers' block in the first place.


```{r Table-Carat-Price, message=FALSE, echo=FALSE, fig.width=4, fig.height=2, fig.cap="Summary of Carat Proportion"}

carat_summary <- diamond %>%
  mutate(carat_bin = cut(carat, breaks = c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 6, 7, 8), right = FALSE)) %>%
  drop_na(carat_bin) %>%
  count(carat_bin) %>%
  mutate(proportions = round(n / sum(n), 3))
kable(carat_summary, col.names = c("Carat Range", "Counts", "Proportion"),
      caption = "Summary of diamond by Carat Range") %>%
  kable_styling(full_width = FALSE, position = "center", latex_options = c("HOLD_position", "striped")) %>%
  column_spec(1, width = "5cm") %>%
  column_spec(2, width = "3cm") %>%
  column_spec(3, width = "3cm")

```


# Bivariate Analysis

## Bivariate Anlaysis on Price

-- General Introduction -- 

## Violin Plots of Price Distributions

```{r violin-plot-full-scale, message = FALSE, echo = FALSE, warning = FALSE, fig.width= 10, fig.height = 5.5, fig.cap = "Violin plots of diamond prices by cut, color, clarity, and carat bins"}
g1 <- ggplot(diamond, aes(x = cut, y = price/10000, fill = cut)) + 
  geom_violin() +
  labs(title = "Diamond Price by Cut",
       x = "Diamond Cut",
       y = "Price (in $10,000s)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8))


g2 <- ggplot(diamond, aes(x = color, y = price/10000, fill = color)) + 
  geom_violin() +
  labs(title = "Diamond Price by Color",
       x = "Diamond Cut",
       y = "Price (in $10,000s)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8))

g3 <- ggplot(diamond, aes(x = clarity, y = price/10000, fill = clarity)) + 
  geom_violin() +
  labs(title = "Diamond Price by Clarity",
       x = "Diamond Clarity",
       y = "Price (in $10,000s)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8))

g4 <- ggplot(diamond, aes(x = carat_bin, y = price/10000, fill = carat_bin)) + 
  geom_violin() +
  labs(title = "Diamond Price by Cut",
       x = "Carat Bin",
       y = "Price (in $10,000s)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8))

(g1 | g2) / (g3 | g4) + 
  plot_annotation(title = "Violin Plot of Diamond Categories", 
  theme = theme(plot.title = element_text(hjust = 0.5)))
```

A random paragraph can also be an excellent way for a writer to tackle writers' block. Writing block can often happen due to being stuck with a current project that the writer is trying to complete. By inserting a completely random paragraph from which to begin, it can take down some of the issues that may have been causing the writers' block in the first place.

```{r violin-plot-zoom-in, message = FALSE, warning = FALSE, echo = FALSE, fig.width= 10, fig.height = 5.5, fig.cap = "Zoomed-in violin plots of diamond prices (\\$0 to \\$10,000) by cut, color, clarity, and carat bins"}
g1 <- ggplot(diamond, aes(x = cut, y = price/10000, fill = cut)) + 
  geom_violin() +
  scale_y_continuous(limits = c(0, 1))  +
  theme_minimal() +
  labs(title = "Diamond Price by Cut",
       x = "Diamond Cut",
       y = "Price (in $10,000s)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8))

g2 <- ggplot(diamond, aes(x = color, y = price/10000, fill = color)) + 
  geom_violin() +
  scale_y_continuous(limits = c(0, 1))  +
  labs(title = "Diamond Price by Color",
       x = "Diamond Cut",
       y = "Price (in $10,000s)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8))

g3 <- ggplot(diamond, aes(x = clarity, y = price/10000, fill = clarity)) + 
  geom_violin() +
  scale_y_continuous(limits = c(0, 1))  +
  labs(title = "Diamond Price by Clarity",
       x = "Diamond Clarity",
       y = "Price (in $10,000s)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8))

g4 <- ggplot(diamond, aes(x = carat_bin, y = price/10000, fill = carat_bin)) + 
  geom_violin() +
  scale_y_continuous(limits = c(0, 1))  +
  labs(title = "Diamond Price by Color",
       x = "Carat Bin",
       y = "Price (in $10,000s)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8))

(g1 | g2) / (g3 | g4) + 
  plot_annotation(title = "Zoom-In Violin Plot of Diamond Categories", 
  theme = theme(plot.title = element_text(hjust = 0.5)))
```



A random paragraph can also be an excellent way for a writer to tackle writers' block. Writing block can often happen due to being stuck with a current project that the writer is trying to complete. By inserting a completely random paragraph from which to begin, it can take down some of the issues that may have been causing the writers' block in the first place.




```{r Table comb short tab, echo=FALSE}
cut_Min_Tab<-CutCarat%>%
  group_by(carat_bin)%>%
  rename(Ave_Price_Cut=Ave_Price)%>%
  mutate(Min_Cut=min(Ave_Price_Cut))%>%
  filter(Ave_Price_Cut==Min_Cut)

cut_Max_Tab<-CutCarat%>%
  group_by(carat_bin)%>%
  rename(Ave_Price_Cut=Ave_Price)%>%
  mutate(Max_Cut=max(Ave_Price_Cut))%>%
  filter(Ave_Price_Cut==Max_Cut)

color_Min_Tab<-ColorCarat%>%
  group_by(carat_bin)%>%
  rename(Ave_Price_Color=Ave_Price)%>%
  mutate(Min_Color=min(Ave_Price_Color))%>%
  filter(Ave_Price_Color==Min_Color)

color_Max_Tab<-ColorCarat%>%
  group_by(carat_bin)%>%
  rename(Ave_Price_Color=Ave_Price)%>%
  mutate(Max_Color=max(Ave_Price_Color))%>%
  filter(Ave_Price_Color==Max_Color)

clarity_Min_Tab<-ClarityCarat%>%
  group_by(carat_bin)%>%
  rename(Ave_Price_Clar=Ave_Price)%>%
  mutate(Min_Clar=min(Ave_Price_Clar))%>%
  filter(Ave_Price_Clar==Min_Clar)

clarity_Max_Tab<-ClarityCarat%>%
  group_by(carat_bin)%>%
  rename(Ave_Price_Clar=Ave_Price)%>%
  mutate(Max_Clar=max(Ave_Price_Clar))%>%
  filter(Ave_Price_Clar==Max_Clar)

clarity_Full_Tab<-bind_rows(clarity_Min_Tab,clarity_Max_Tab)%>%
  arrange(carat_bin,Ave_Price_Clar)

color_Full_Tab<-bind_rows(color_Min_Tab,color_Max_Tab)%>%
  arrange(carat_bin,Ave_Price_Color)

cut_Full_Tab<-bind_rows(cut_Min_Tab,cut_Max_Tab)%>%
  arrange(carat_bin,Ave_Price_Cut)

clarity_tab_calc<-clarity_Full_Tab%>%
  group_by(carat_bin)%>%
  summarize(Price_Rng_Clar=max(Ave_Price_Clar)-min(Ave_Price_Clar))%>%
  dplyr::select(Price_Rng_Clar)

color_tab_calc<-color_Full_Tab%>%
  group_by(carat_bin)%>%
  summarize(Price_Rng_Color=max(Ave_Price_Color)-min(Ave_Price_Color))%>%
  dplyr::select(Price_Rng_Color)

cut_tab_calc<-cut_Full_Tab%>%
  group_by(carat_bin)%>%
  summarize(Price_Rng_Cut=max(Ave_Price_Cut)-min(Ave_Price_Cut))%>%
  dplyr::select(carat_bin,Price_Rng_Cut)

Full_Tab<-bind_cols(cut_tab_calc,color_tab_calc,clarity_tab_calc)
```



## Carat vs Clarity: Proportion and Price Analysis


```{r clarity-and-carat-heatmap, warning=FALSE, message=FALSE, echo = FALSE, results="hide", error=FALSE, fig.width=6, fig.height=4.5, fig.align="center", fig.cap = "Diamond Carat Bin vs Clarity: Proportion and Price Heatmaps"}
# new dataframe with carat new bins and clarity ordered
# new dataframe with carat new bins and clarity ordered
carat_new_bins <- clarity_ordered %>%
  drop_na(carat) %>%
  mutate(carat_bin = cut(carat,
                         breaks = c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 8),
                         labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1", "1-1.5", "1.5-2", "2-3", "3-4", "4-5", "5-8"),
                         right = FALSE)) %>%
  drop_na(carat_bin)

# compare carat and clarity by count 
carat_clarity1<-carat_new_bins%>%
  count(carat_bin,clarity)%>%
  group_by(carat_bin)%>%
  mutate(Percentage= round(n/sum(n)*100,2))

# compare carat and clarity by price
carat_clarity2<-carat_new_bins%>%
  group_by(clarity,carat_bin)%>%
  summarise(avg_price=round(mean(price),0))%>%
  ungroup()
 
# Heat maps for diamond carat by clarity
a1 <-ggplot(carat_clarity1,aes(x=carat_bin,y=clarity,fill=Percentage))+
  geom_tile()+
  scale_fill_gradient(low="white",high="red") +
  geom_text(aes(label = Percentage), size = 2.5)+
  labs(x="Carat Bin",
       y="Clarity",
       title="Quantity",
       fill = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

a2 <- ggplot(carat_clarity2,aes(x=carat_bin,y=clarity,fill=avg_price/1000))+
  geom_tile()+
  scale_fill_gradient(low="white",high="red")+
  geom_text(aes(label = round(avg_price/1000,2)), size = 2.5) +
  labs(x="Carat Bin",y="Clarity",
       title="Average Price ($1,000s)",
       fill = "Avg. Price ($1,000s)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

a1/a2 + 
  plot_annotation(title = "Diamond Carat Bin vs Clarity Overview",
                  subtitle = "Proportion and Average Price Heatmaps",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16),
                    plot.subtitle = element_text(hjust = 0.5, size =8 )
  ))
```

## Carat vs Cut: Proportion and Price Analysis

```{r carat-cut-heatmap, warning=FALSE, message=FALSE, echo = FALSE, results="hide", error=FALSE, fig.width=6, fig.height=4.5, fig.align="center", fig.cap="Diamond Carat Bin vs Cut: Proportion and Price Heatmaps"}
b1 <- ggplot(CutCarat,aes(x=carat_bin,y=cut,fill=Ave_Price/1000))+
  geom_tile()+
  scale_fill_gradient(low="white",high="blue")+
  geom_text(aes(label=round(Ave_Price/1000, 2)), size = 2.5)+
  labs(x="Carat Bin",
       y="Cut",
       title="Average Price ($1,000s)",
       fill = "Avg. Price ($1,000s)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

cut_carrat <- diamond %>%
  count(cut, carat_bin) %>%
  group_by(cut) %>%
  mutate(Percentage= round(n/sum(n)*100,2))

  
b2 <- ggplot(cut_carrat, aes(x = carat_bin, y = cut, fill = Percentage))+
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  geom_text(aes(label = Percentage), size = 2.5) +
  labs(
    x = "Carat Bin",
    y = "Cut",
    title = "Quantity",
    fill = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))


b2/b1 + 
  plot_annotation(title = "Diamond Carat vs Cut Overview",
                  subtitle = "Proportion and Average Price Heatmaps",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16),
                    plot.subtitle = element_text(hjust = 0.5, size =8 )
  ))
```

## Color vs Carat: Proportion and Price Analysis

```{r color-carat-heatmap, warning=FALSE, message=FALSE, echo = FALSE, results="hide", error=FALSE, fig.width=6, fig.height=4.5, fig.align="center", fig.cap="Diamond Color vs Carat Bin: Proportion and Price Heatmaps"}
c1 <- ggplot(ColorCarat,aes(x=color,y=carat_bin,fill=Ave_Price/1000))+
  geom_tile()+
  scale_fill_gradient(low="white",high="darkgreen")+
  geom_text(aes(label=round(Ave_Price/1000, 2)), size = 2.5)+
  labs(x="Color",
       y="Carat Bin",
       title="Average Price ($1,000s)",
       fill = "Avg. Price ($1,000s)") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

color_carat <- diamond %>%
  count(color, carat_bin) %>%
  group_by(color) %>%
  mutate(Percentage= round(n/sum(n)*100,2))

  
c2 <- ggplot(color_carat, aes(x = color, y = carat_bin, fill = Percentage))+
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  geom_text(aes(label = Percentage), size = 2.5) +
  labs(
    x = "Color",
    y = "Carat Bin",
    title = "Quantity",
    fill = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

c2/c1 + 
  plot_annotation(title = "Diamond Color vs Carat Bin Overview",
                  subtitle = "Proportion and Average Price Heatmaps",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16),
                    plot.subtitle = element_text(hjust = 0.5, size =8 )
  ))
```


## Cut vs Color: Proportion and Price Analysis

```{r color-cut-heatmap, warning=FALSE, message=FALSE, echo = FALSE, results="hide", fig.width=6, fig.height=4.5, fig.align="center", fig.cap="Diamond Cut vs Color: Proportion and Price Heatmaps"}
color_cut <- diamond %>%
  count(cut,color) %>%
  group_by(color) %>%
  mutate(Percentage= round(n/sum(n)*100,2))


d1 <- ggplot(color_cut,aes(x=cut, y=color, fill=Percentage))+
  geom_tile()+
  scale_fill_gradient(low="white",high="purple")+
  geom_text(aes(label=Percentage), size = 2.5)+
  labs(
    x = "Cut",
    y = "Color",
    title = "Quantity",
    fill = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

d2 <- ggplot(CutColor, aes(x=cut,y=color,fill=Ave_Price/1000))+
  geom_tile()+
  scale_fill_gradient(low="white",high="purple")+
  geom_text(aes(label=round(Ave_Price/1000,2)), size = 2.5)+
  labs(x="Cut",
       y="Color",
       title="Average Price ($1,000s)",
       fill = "Avg. Price ($1,000s)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

d1/d2 + 
  plot_annotation(title = "Diamond Cut vs Color Overview",
                  subtitle = "Proportion and Average Price Heatmaps",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16),
                    plot.subtitle = element_text(hjust = 0.5, size =8 )
  ))
```

## Color vs Clarity: Proportion and Price Analysis

```{r color-clarity-heatmap, warning=FALSE, message=FALSE, echo = FALSE, results="hide", fig.width=6, fig.height=4.5, fig.align="center", fig.cap="Diamond Clarity vs Color: Proportion and Price Heatmaps"}
color_clarity <- diamond %>%
  count(color, clarity) %>%
  group_by(color) %>%
  mutate(Percentage= round(n/sum(n)*100,2))


e1 <- ggplot(color_clarity, aes(x = color, y = clarity, fill = Percentage)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "brown4") +
  geom_text(aes(label = Percentage), size = 2.5) +
  labs(
    x = "Color",
    y = "Clarity",
    title = "Quantity",
    fill = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

# compare color and clarity by count 
color_clarity1 <- clarity_ordered %>%
  count(color, clarity) %>%
  group_by(color) %>%
  mutate(Percentage= round(n/sum(n)*100,2))

# compare color and clarity by price 
color_clarity2 <- clarity_ordered%>%
  group_by(color,clarity)%>%
  summarise(avg_price=round(mean(price),0))%>%
  ungroup()


e2 <- ggplot(color_clarity2,aes(x=color,y=clarity,fill=avg_price))+
  geom_tile()+
  scale_fill_gradient(low="white",high="brown4")+
  geom_text(aes(label=round(avg_price/1000, 2)), size = 2.5) +
  labs(x="Color",
       y="Clarity",
       title="Average Price ($1,000s)",
       fill = "Avg. Price ($1,000s)") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))

e1/e2 + 
  plot_annotation(title = "Diamond Clarity vs Color Overview",
                  subtitle = "Proportion and Average Price Heatmaps",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16),
                    plot.subtitle = element_text(hjust = 0.5, size =8 )
  ))
```

## Clarity vs Color: Proportion and Price Analysis


```{r cut-clarity-heatmap, warning=FALSE, message=FALSE, echo = FALSE, results="hide", fig.width=6, fig.height=4.5, fig.align="center", fig.cap="Diamond Clarity vs Color: Proportion and Price Heatmaps"}
clarity_cut <- diamond %>%
  count(cut, clarity) %>%
  group_by(cut) %>%
  mutate(Percentage= round(n/sum(n)*100,2))

  
f1 <- ggplot(clarity_cut, aes(x = cut, y = clarity, fill = Percentage))+
  geom_tile() +
  scale_fill_gradient(low = "white", high = "orange3") +
  geom_text(aes(label = Percentage), size = 2.5) +
  labs(
    x = "Cut",
    y = "Clarity",
    title = "Quantity",
    fill = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))


# new dataframe with cut and clarity ordered 
cut_clarity<-diamond%>%
  mutate(cut=cut%>%
           fct_relevel(c("Good","Very Good","Ideal","Astor Ideal")))%>%
  mutate(clarity=clarity%>%
           fct_relevel(c("SI2","SI1","VS2","VS1","VVS2","VVS1","IF","FL")))

# compare cut and clarity by count 
cut_clarity1<-cut_clarity%>%
  count(cut,clarity)%>%
  group_by(cut)%>%
  mutate(Percentage= round(n/sum(n)*100,2))

# compare cut and clarity by price
cut_clarity2<-cut_clarity%>%
  group_by(cut,clarity)%>%
  summarise(avg_price=round(mean(price),0))%>%
  ungroup()
 

f2 <- ggplot(cut_clarity2,aes(x=cut,y=clarity,fill=avg_price/1000))+
  geom_tile()+
  scale_fill_gradient(low="white",high="orange3") +
  geom_text(aes(label = round(avg_price/1000, 2)), size = 2.5) +
  labs(x="Cut",
       y="Clarity",
       title="Average Price ($1,000s)",
       fill = "Avg. Price ($1,000s)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust=0.5, size = 12),
        legend.title = element_text(hjust=0.5, size = 11))


f1/f2 + 
  plot_annotation(title = "Diamond Cut vs Clarity Overview",
                  subtitle = "Proportion and Average Price Heatmaps",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, size = 16),
                    plot.subtitle = element_text(hjust = 0.5, size =8 )
  ))
```





# Simple Linear Regression Modeling: Diamond Price vs Carat

```{r Scatter-Price,message=FALSE,echo=FALSE, fig.width=4, fig.height=3, fig.cap="Diamond Price vs Carat Size",fig.align="center"}
ggplot(diamond, aes(x = carat, y = price)) +
  geom_point(color = "skyblue", alpha = 1) +
  labs(title = "Diamond Price vs Carat",x = "Carat",y = "Price") +
  geom_smooth(method = "lm", color = "blue") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
  
```

```{r Table-Carat-Price-Var, warning=FALSE, message=FALSE, echo = FALSE, fig.width=5, fig.height=3,fig.align="center",fig.cap="Carat Price Variability"}
carat_price_summary2 <- diamond %>%
  mutate(carat_bin = cut(carat, breaks = c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 6, 7, 8), right = FALSE)) %>%
  group_by(carat_bin) %>%
  summarise(
    Min_Price = min(price, na.rm = TRUE),
    Max_Price = max(price, na.rm = TRUE),
    Avg_Price = round(mean(price, na.rm = TRUE), 2),
    Variability_Percent = round((Max_Price - Min_Price) / Min_Price * 100,2))
kable(carat_price_summary2, col.names = c("Carat Range", "Min Price", "Max Price", "Average Price", "Variability (%)"),
      caption = "Summary of Diamond Prices by Carat Range") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(1, width = "5cm") %>%
  column_spec(2, width = "3cm") %>%
  column_spec(3, width = "3cm")
```



```{r Residual-Plot, message=FALSE, echo=FALSE,fig.width=7, fig.height=4,fig.align="center",fig.cap="Residuals plot"}
model_carat <- lm(price ~ carat, diamond)
par(mfrow = c(2,2))
plot(model_carat)
residuals_model <- residuals(model_carat)
```

```{r ACF-plot, message=FALSE, echo=FALSE,fig.width=4, fig.height=3,fig.align="center",fig.cap="ACF Residual Model"}
residuals_model <- residuals(model_carat)
acf(residuals_model)
```

```{r Box-Cox, message=FALSE, echo=FALSE,fig.width=4, fig.height=3,fig.align="center",fig.cap="Box-Cox Transformation Plot for the Diamond Price vs Carat Model"}
boxcox(model_carat)
```

```{r Residual-Plot-Trans, message=FALSE, fig.width=7, fig.height=4,echo=FALSE,fig.align="center",fig.cap="Residuals plot"}
diamond$price_sqrt <- sqrt(diamond$price)
model_carat_sqrt <- lm(price_sqrt ~ carat, diamond)


par(mfrow = c(2, 2))
plot(model_carat_sqrt)
```



```{r ACF-plot-tran, message=FALSE, echo=FALSE,fig.width=4, fig.height=3,fig.align="center",fig.cap="ACF Residual Model after Square Root Transformation "}
residuals_model_sq <- residuals(model_carat_sqrt)
acf(residuals_model_sq)
```

```{r Summary, warning=FALSE, message=FALSE, echo = FALSE, fig.align='center', fig.height=3, fig.width = 3}
print(summary(model_carat_sqrt))
```

```{r Scatter-Price-post,message=FALSE,echo=FALSE, fig.width=4, fig.height=3, fig.cap="Scatterplot of Carat vs Square Root of Price (after square root transformation) with fitted linear model",fig.align="center"}
ggplot(diamond, aes(x = carat, y = price_sqrt)) +
  geom_point(color = "skyblue", alpha = 1) +
  labs(title = "Diamond Carat vs Square Root of Price",
       x = "Carat",
       y = "Square Root of Price") +
  geom_smooth(method = "lm", color = "blue") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```